\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2024-07-08 scratch/notes_john/JSS-article/JSS-article-3.Rmd
some changes to John's JSS draft

The \pkg{cv} package for \proglang{R} automates the process of cross-validation for standard \proglang{R} statistical model objects. The principal function in the packages, also named \code{cv()}, has methods for objects produced by a number of commonly employed regression-modeling functions, including those for mixed-effects models:
```{r cv-methods}
library("cv", quietly=TRUE)
methods("cv")
```
* The \code{"lm"} and \code{"glm"} methods are for linear and generalized-linear models fit respectively by the standard \proglang{R} \code{lm()} and \code{glm()} functions.

* The \code{"modList"} method for \code{cv()} cross-validates several competing models, not necessarily of the same class, using the same division of the data into folds. The \code{cv()} function is introduced in the context of a preliminary example in Section \ref{preliminary-example-polynomial-regression} of the paper.

* Cross-validating mixed-effects models, using the \code{"glmmTMB"}, \code{"lme"}, and  \code{"merMod"} methods for \code{cv()}, involves special considerations that we take up in Section \ref{cross-validating-mixed-effects-models}.

* The \code{"function"} method for \code{cv()}, discussed in Section \ref{cross-validating-model-specification}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors, or model selection via CV itself.

* The \code{"default"} \code{cv()} method works (perhaps with a bit of coaxing) with many other existing regression-model classes for which there is an \code{update()} method that accepts a \code{data} argument. More generally, the \pkg{cv} package is designed to be extensible, as discussed in Section \ref{extending-the-cv-package}.

@@

The data for the example in this section are drawn from the \pkg{ISLR2} package for \proglang{R}, associated with @JamesEtAl:2021. The presentation here is close (though not identical) to that in the original source [@JamesEtAl:2021 Secs. 5.1, 5.3], and it demonstrates the use of the `cv()` function.[^boot]

[^boot]: @JamesEtAl:2021 use the `cv.glm()` function in the \pkg{boot} package [@CantyRipley2022; @DavisonHinkley:1997]. Despite its name, `cv.glm()` is an independent function and not a method of a `cv()` generic function. The \pkg{boot} package is part of the standard \proglang{R} distribution.

@@

```{r Auto}
data("Auto", package="ISLR2")
summary(Auto)
```
With the exception of `origin` (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see `help("Auto", package="ISLR2")`.

```{r mpg-horsepower-scatterplot-polynomials, echo=FALSE}
#| out.width = "50%",
#| fig.height = 5,
#| fig.cap = "\\code{mpg} vs. \\code{horsepower} for the \\code{Auto} data, showing fitted polynomials of degree 1 through 5."
plot(mpg ~ horsepower, data=Auto)
horsepower <- with(Auto, 
                   seq(min(horsepower), max(horsepower), 
                       length=1000))
for (p in 1:5){
  m <- lm(mpg ~ poly(horsepower,p), data=Auto)
  mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
  lines(horsepower, mpg, col=p + 1, lty=p, lwd=4)
}
legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=4,
       title="Degree", inset=0.02)
```

We'll focus here on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure \ref{fig:mpg-horsepower-scatterplot-polynomials}. The relationship between the two variables is monotone, decreasing, and nonlinear. Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-scatterplot-polynomials}. The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high `mpg` values at the right (but see the CV results reported below).

[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here. Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader). We revisit the \code{Auto} data in Section \ref{cross-validating-model-specification}.

```{r mpg-horsepower-MSE-se, echo=FALSE}
#| out.width = "50%",
#| fig.height = 5,
#| fig.cap = "Estimated squared error as a function of polynomial degree, $p$"

@@

The generic `cv()` function has an `"lm"` method,
```{r cv-lm-method}
args(cv:::cv.lm)
```
which takes the following arguments:

* \code{model}, an \code{"lm"} object, the only required argument.

* \code{data}, which can usually be inferred from the \code{model} object.

* \code{criterion}, a function to compute the CV criterion (defaulting to \code{mse}).

* \code{k}, the number of folds to employ (defaulting to \code{10}); the character value \code{"n"} or \code{"loo"} may be supplied to specify leave-one-out cross-validation.

* \code{reps}, the number of times to repeat the CV procedure (defaulting to 1).

* \code{seed}, the seed for \proglang{R}'s pseudo-random number generator; if not specified a value is randomly selected, reported, and saved, so that the CV procedure is replicable.

* \code{confint}, whether or not to compute a confidence interval for the CV criterion, defaulting to \code{TRUE} if there are at least 400 cases; a confidence interval is computed only if the CV criterion can be expressed as the average of casewise components (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals} for details).

* \code{level}, the level for the confidence interval (defaulting to \code{0.95}).

* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).

* \code{ncores}, the number of cores to employ for parallel computation; if \code{cores = 1} (the default), the computations are not parallelized.

To illustrate, we perform 10-fold CV for a quadratic polynomial fit to the \code{Auto} data:
```{r cv-lm-1}
m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
summary(m.auto)

cv(m.auto, confint=TRUE)
```
The function reports the CV estimate of MSE, a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression. Because the number of cases $n = 392 < 400$ for the \code{Auto} data, we set the argument \code{confint=TRUE} to obtain a confidence interval for the MSE, which proves to be quite wide.

@@

```{r cv.lm-2`}
cv(m.auto, k="loo")
```
The \code{"hatvalues"} method reports only the CV estimate of MSE. Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted. All three methods produce exact results for a linear model (within the precision of floating-point computations):
```{r cv.lm-3}
cv(m.auto, k="loo", method="naive", confint=TRUE)

cv(m.auto, k="loo", method="Woodbury", confint=TRUE)
```
The \code{"naive"} and \code{"Woodbury"} methods also return the bias-adjusted estimate of MSE (and a confidence interval around it) along with the full-sample MSE, but bias isn't an issue for LOO CV.

This is a small regression problem and all three computational approaches are essentially instantaneous, but it is still of interest to investigate their relative speed. In the following comparison, we include the \code{cv.glm()} function from the \pkg{boot} package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM. We use the \code{microbenchmark()} function from the package of the same name  [@Mersmann:2023] for the timings:
```{r cv.lm.timings, cache=TRUE}
m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
boot::cv.glm(Auto, m.auto.glm)$delta # MSE, biased-corrected MSE

microbenchmark::microbenchmark(
  hatvalues = cv(m.auto, k="loo"),
  Woodbury = cv(m.auto, k="loo", method="Woodbury"),
  naive = cv(m.auto, k="loo", method="naive"),
  cv.glm = boot::cv.glm(Auto, m.auto.glm),
  times=10
)
```
On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.

@@

We illustrate with the polynomial regression models of varying degree for the \code{Auto} data, beginning by fitting and saving the 10 models:
```{r polyomial-models}
mlist <- vector(10, mode="list")
for (p in 1:10) mlist[[p]] <- lm(mpg ~ poly(horsepower, p), data = Auto)
names(mlist) <- paste0("m.", 1:10)
mlist[2] # e.g., the quadratic fit
```
We then apply \code{cv()} to the list of 10 models (the \code{data} argument is required):
```{r polynomial-regression-CV}
# 10-fold CV
mlist <- models(mlist) # create "modList" object

@@

# LOO CV
cv.auto.loo <- cv(mlist, data=Auto, k="loo")
cv.auto.loo[2] # e.g., for quadratic model
```
The \code{models()} function takes an arbitrary number of regression models as its arguments, which are optionally named, to create a \code{"modList"} object. Alternatively, as here, we can supply a pre-constructed, optionally named, list of models as the single argument to \code{models()}. To visualize the results, comparing the models, we invoke the \code{plot()} method for the \code{"cvModList"} object returned by \code{cv()} (see Figure \ref{fig:polynomial-regression-CV-graph-2}):

@@

In contrast, mixed-effects models are fit to *dependent* data, in which cases are clustered, such as hierarchical data, where the clusters comprise higher-level units (e.g., students clustered in schools), or longitudinal data, where the clusters are individuals and the cases are repeated observations on the individuals over time.[^crossed-effects]

[^crossed-effects]: There are, however, more complex situations that give rise to so-called *crossed* (rather than *nested*) random effects. For example, consider students within classes within schools. In primary schools, students typically are in a single class, and so classes are nested within schools. In secondary schools, however, students typically take several classes and students who are together in a particular class may not be together in other classes; consequently, random effects based on classes within schools are crossed. The \code{lmer()} function in the \pkg{lme4} package, for example, is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects. We present an example of the latter in a vignette for the \pkg{cv} package.

We can think of two approaches to applying cross-validation to clustered data:[^cv-faq]

[^cv-faq]: We subsequently discovered that @Vehtari:2023 [Section 8] makes similar points.

@@

The HSB data are included in the \code{MathAchieve} and \code{MathAchSchool} data sets in the \pkg{nlme} package  [@PinheiroBates:2000]. \code{MathAchieve} comprises individual-level data on 7185  students in 160 high schools, and \code{MathAchSchool} contains school-level data:
```{r HSB-data}
data("MathAchieve", package="nlme")
dim(MathAchieve)
head(MathAchieve, 3)
tail(MathAchieve, 3)

data("MathAchSchool", package="nlme")
dim(MathAchSchool)
head(MathAchSchool, 2)
tail(MathAchSchool, 2)
```
The first few students are in school number 1224 and the last few in school 9586.

@@

Some data-management is required before fitting a mixed-effects model to the HSB data:
```{r HSB-data-management, cache=TRUE}
HSB <- MathAchieve
HSB <- merge(MathAchSchool[, c("School", "Sector")],
             HSB[, c("School", "SES", "MathAch")], by="School")
names(HSB) <- tolower(names(HSB))
HSB <- within(HSB, {
  mean.ses <- ave(ses, school)
  cses <- ses - mean.ses
})
```
In the process, we merge variables from the school-level and student-level data sets, and create two new school-level variables: \code{mean.ses}, which is the average SES for students in each school; and \code{cses}, which is the individual students' SES centered at their school means. For details, see @FoxWeisberg:2019 [Sec. 7.2.2].

Still following Fox and Weisberg, we proceed to use the \code{lmer()} function in the \pkg{lme4} package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data:
```{r HSB-lmer, cache=TRUE}
library("lme4", quietly=TRUE)
hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
                   + (cses | school), data=HSB)
summary(hsb.lmer, correlation=FALSE)
```

We can then cross-validate at the cluster (i.e., school) level,
```{r HSB-lmer-CV-cluster, cache=TRUE}
cv(hsb.lmer, k=10, clusterVariables="school", seed=5240)
```
or at the case (i.e., student) level,
```{r HSB-lmer-CV-case, cache=TRUE}
cv(hsb.lmer, seed=1575)
```
For cluster-level CV, the \code{clusterVariables} argument tells \code{cv()} how the clusters are defined. Were there more than one clustering variable, say classes within schools, these would be provided as a character vector of variable names: \code{clusterVariables = c("school", "class")}. For cluster-level CV, the default is \code{k = "loo"}, that is, leave one cluster out at a time; we instead specify \code{k = 10} folds of clusters, each fold therefore comprising $160/10 = 16$ schools.

@@

We start by generating the data, using parameters consistent with the description above and meant to highlight the issues that arise in cross-validating mixed-effects models:[^1]

[^1]: We invite the interested reader to experiment with varying the parameters of our example.

@@

We proceed to fit several mixed-effects models to the data, using the `compareCoefs()` function from the \pkg{car} package [@FoxWeisberg:2019] to display the fixed-effect estimates for these models. We discuss each of these models below.
```{r fitted-mixed-models}
mod.0 <- lmer(y ~ 1 + (1 | group), Data)
mod.1 <- lmer(y ~ x + (1 | group), Data)
mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
library("car", quietly=TRUE)
compareCoefs(mod.0, mod.1, mod.2, mod.3)
```

@@

To illustrate cross-validation of model specification, we return to the \code{Auto} data set:[^Venables]
```{r Auto-redux}
names(Auto)
xtabs(~ year, data=Auto)
xtabs(~ origin, data=Auto)
xtabs(~ cylinders, data=Auto)
```

[^Venables]: This example benefits from an email conversation with Bill Venables, who of course isn't responsible for the use to which we've put his insightful remarks.

The \code{Auto} data appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}. Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data. We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore}, produced by the \code{scatterplotMatrix()} function in the \pkg{car} package):
```{r Auto-explore}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Scatterplot matrix for the numeric variables in the \\code{Auto} data. In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression."
Auto$cylinders <- factor(Auto$cylinders,
                         labels=c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels=c("America", "Europe", "Japan"))
rownames(Auto) <- make.names(Auto$name, unique=TRUE)
Auto$name <- NULL

scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
                  smooth=list(spread=FALSE), data=Auto, pch=".")
```
A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors. There are, however, only 5 distinct values of \code{cylinders} (ranging from 3 to 8), but cars with 3 or 5 cylinders are rare, and none of the cars has 7 cylinders. There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^year] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.

[^year]: Making the decision to treat \code{year} as a factor on this basis could be construed as cheating in the current context, which illustrates the difficulty of automating the whole model-selection process. It's rarely desirable, in our opinion, to forgo exploration of the data to ensure the purity of model validation. We believe, however, that it's still useful to automate as much of the process as we can, as illustrated here, to obtain a more realistic, if still biased, estimate of the predictive power of a model.

We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
```{r Auto-working-model}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Component+residual plots for the working model fit to the \\code{Auto} data. The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths."
m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch=".")
```
The component+residual plots in Figure \ref{fig:Auto-working-model} clearly reveal the inadequacy of the model.

@@

Applying `selectModelList()` to the `Auto` polynomial-regression models, and using 10-fold CV, we obtain:
```{r recursive-CV-polynomials}
recursiveCV.auto <- cv(selectModelList, data=Auto, 
                       working.model=mlist,
                       save.model=TRUE, seed=2120)
recursiveCV.auto
recursiveCV.auto$selected.model
cv(mlist[[7]], seed=2120) # same seed for same folds
```
As expected, recursive CV produces a larger estimate of MSE for the selected 7th degree polynomial model than CV applied directly to this model.

>>>>

The \pkg{cv} package for \proglang{R} automates the process of cross-validation for standard \proglang{R} statistical model objects. The principal function in the packages, also named \code{cv()}, has methods for objects produced by a number of commonly employed regression-modeling functions, including those for mixed-effects models:

```{r cv-methods}
library("cv", quietly=TRUE)
methods("cv")
```

-   The \code{"lm"} and \code{"glm"} methods are for linear and generalized-linear models fit respectively by the standard \proglang{R} \code{lm()} and \code{glm()} functions.

-   The \code{"modList"} method for \code{cv()} cross-validates several competing models, not necessarily of the same class, using the same division of the data into folds. The \code{cv()} function is introduced in the context of a preliminary example in Section \ref{preliminary-example-polynomial-regression} of the paper.

-   Cross-validating mixed-effects models, using the \code{"glmmTMB"}, \code{"lme"}, and \code{"merMod"} methods for \code{cv()}, involves special considerations that we take up in Section \ref{cross-validating-mixed-effects-models}.

-   The \code{"function"} method for \code{cv()}, discussed in Section \ref{cross-validating-model-specification}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors, or model selection via CV itself.

-   The \code{"default"} \code{cv()} method works (perhaps with a bit of coaxing) with many other existing regression-model classes for which there is an \code{update()} method that accepts a \code{data} argument. More generally, the \pkg{cv} package is designed to be extensible, as discussed in Section \ref{extending-the-cv-package}.

@@

The data for the example in this section are drawn from the \pkg{ISLR2} package for \proglang{R}, associated with @JamesEtAl:2021. The presentation here is close (though not identical) to that in the original source [@JamesEtAl:2021 Secs. 5.1, 5.3], and it demonstrates the use of the `cv()` function.[^1]

[^1]: @JamesEtAl:2021 use the `cv.glm()` function in the \pkg{boot} package [@CantyRipley2022; @DavisonHinkley:1997]. Despite its name, `cv.glm()` is an independent function and not a method of a `cv()` generic function. The \pkg{boot} package is part of the standard \proglang{R} distribution.

@@

```{r Auto}
data("Auto", package="ISLR2")
summary(Auto)
```

With the exception of `origin` (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see `help("Auto", package="ISLR2")`.

```{r mpg-horsepower-polynomials, fig.show="hold", echo=FALSE}
#| out.width = "45%",
#| fig.height = 5,
#| fig.cap = "(a) \\code{mpg} vs. \\code{horsepower} for the \\code{Auto} data, showing fitted polynomials of degree 1 through 5. (b) Estimated squared error as a function of polynomial degree, $p$.",
#| fig.subcap=rep("", 2)
plot(mpg ~ horsepower, data=Auto)
horsepower <- with(Auto, 
                   seq(min(horsepower), max(horsepower), 
                       length=1000))
for (p in 1:5){
  m <- lm(mpg ~ poly(horsepower,p), data=Auto)
  mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
  lines(horsepower, mpg, col=p + 1, lty=p, lwd=4)
}
legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=4,
       title="Degree", inset=0.02)

@@

We'll focus here on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure \ref{fig:mpg-horsepower-scatterplot-polynomials}. The relationship between the two variables is monotone, decreasing, and nonlinear. Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^2] Polynomial fits for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-scatterplot-polynomials}. The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high `mpg` values at the right (but see the CV results reported below).

[^2]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here. Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader). We revisit the \code{Auto} data in Section \ref{cross-validating-model-specification}.

The generic `cv()` function has an `"lm"` method,

```{r cv-lm-method}
args(cv:::cv.lm)
```

which takes the following arguments:

-   \code{model}, an \code{"lm"} object, the only required argument.

-   \code{data}, which can usually be inferred from the \code{model} object.

-   \code{criterion}, a function to compute the CV criterion (defaulting to \code{mse}).

-   \code{k}, the number of folds to employ (defaulting to \code{10}); the character value \code{"n"} or \code{"loo"} may be supplied to specify leave-one-out cross-validation.

-   \code{reps}, the number of times to repeat the CV procedure (defaulting to 1).

-   \code{seed}, the seed for \proglang{R}'s pseudo-random number generator; if not specified a value is randomly selected, reported, and saved, so that the CV procedure is replicable.

-   \code{confint}, whether or not to compute a confidence interval for the CV criterion, defaulting to \code{TRUE} if there are at least 400 cases; a confidence interval is computed only if the CV criterion can be expressed as the average of casewise components (see Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals} for details).

-   \code{level}, the level for the confidence interval (defaulting to \code{0.95}).

-   \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).

-   \code{ncores}, the number of cores to employ for parallel computation; if \code{cores = 1} (the default), the computations are not parallelized.

To illustrate, we perform 10-fold CV for a quadratic polynomial fit to the \code{Auto} data:

```{r cv-lm-1}
m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
summary(m.auto)

cv(m.auto, confint=TRUE)
```

The function reports the CV estimate of MSE, a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression. Because the number of cases $n = 392 < 400$ for the \code{Auto} data, we set the argument \code{confint=TRUE} to obtain a confidence interval for the MSE, which proves to be quite wide.

@@

```{r cv.lm-2`}
cv(m.auto, k="loo")
```

The \code{"hatvalues"} method reports only the CV estimate of MSE. Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted. All three methods produce exact results for a linear model (within the precision of floating-point computations):

```{r cv.lm-3}
cv(m.auto, k="loo", method="naive", confint=TRUE)

cv(m.auto, k="loo", method="Woodbury", confint=TRUE)
```

The \code{"naive"} and \code{"Woodbury"} methods also return the bias-adjusted estimate of MSE (and a confidence interval around it) along with the full-sample MSE, but bias isn't an issue for LOO CV.

This is a small regression problem and all three computational approaches are essentially instantaneous, but it is still of interest to investigate their relative speed. In the following comparison, we include the \code{cv.glm()} function from the \pkg{boot} package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM. We use the \code{microbenchmark()} function from the package of the same name [@Mersmann:2023] for the timings:

```{r cv.lm.timings, cache=TRUE}
m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
boot::cv.glm(Auto, m.auto.glm)$delta # MSE, biased-corrected MSE

microbenchmark::microbenchmark(
  hatvalues = cv(m.auto, k="loo"),
  Woodbury = cv(m.auto, k="loo", method="Woodbury"),
  naive = cv(m.auto, k="loo", method="naive"),
  cv.glm = boot::cv.glm(Auto, m.auto.glm),
  times=10
)
```

On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.

@@

We illustrate with the polynomial regression models of varying degree for the \code{Auto} data, beginning by fitting and saving the 10 models:

```{r polyomial-models}
mlist <- vector(10, mode="list")
for (p in 1:10) mlist[[p]] <- lm(mpg ~ poly(horsepower, p), data = Auto)
names(mlist) <- paste0("m.", 1:10)
mlist[2] # e.g., the quadratic fit
```

We then apply \code{cv()} to the list of 10 models (the \code{data} argument is required):

```{r polynomial-regression-CV}
# 10-fold CV
mlist <- models(mlist) # create "modList" object

@@

# LOO CV
cv.auto.loo <- cv(mlist, data=Auto, k="loo")
cv.auto.loo[2] # e.g., for quadratic model
```

The \code{models()} function takes an arbitrary number of regression models as its arguments, which are optionally named, to create a \code{"modList"} object. Alternatively, as here, we can supply a pre-constructed, optionally named, list of models as the single argument to \code{models()}. To visualize the results, comparing the models, we invoke the \code{plot()} method for the \code{"cvModList"} object returned by \code{cv()} (see Figure \ref{fig:polynomial-regression-CV-graph-2}):

@@

In contrast, mixed-effects models are fit to *dependent* data, in which cases are clustered, such as hierarchical data, where the clusters comprise higher-level units (e.g., students clustered in schools), or longitudinal data, where the clusters are individuals and the cases are repeated observations on the individuals over time.[^3]

[^3]: There are, however, more complex situations that give rise to so-called *crossed* (rather than *nested*) random effects. For example, consider students within classes within schools. In primary schools, students typically are in a single class, and so classes are nested within schools. In secondary schools, however, students typically take several classes and students who are together in a particular class may not be together in other classes; consequently, random effects based on classes within schools are crossed. The \code{lmer()} function in the \pkg{lme4} package, for example, is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects. We present an example of the latter in a vignette for the \pkg{cv} package.

We can think of two approaches to applying cross-validation to clustered data:[^4]

[^4]: We subsequently discovered that @Vehtari:2023 [Section 8] makes similar points.

@@

The HSB data are included in the \code{MathAchieve} and \code{MathAchSchool} data sets in the \pkg{nlme} package [@PinheiroBates:2000]. \code{MathAchieve} comprises individual-level data on 7185 students in 160 high schools, and \code{MathAchSchool} contains school-level data:

```{r HSB-data}
data("MathAchieve", package="nlme")
dim(MathAchieve)
head(MathAchieve, 3)
tail(MathAchieve, 3)

data("MathAchSchool", package="nlme")
dim(MathAchSchool)
head(MathAchSchool, 2)
tail(MathAchSchool, 2)
```

The first few students are in school number 1224 and the last few in school 9586.

@@

Some data-management is required before fitting a mixed-effects model to the HSB data:

```{r HSB-data-management, cache=TRUE}
HSB <- MathAchieve
HSB <- merge(MathAchSchool[, c("School", "Sector")],
             HSB[, c("School", "SES", "MathAch")], by="School")
names(HSB) <- tolower(names(HSB))
HSB <- within(HSB, {
  mean.ses <- ave(ses, school)
  cses <- ses - mean.ses
})
```

In the process, we merge variables from the school-level and student-level data sets, and create two new school-level variables: \code{mean.ses}, which is the average SES for students in each school; and \code{cses}, which is the individual students' SES centered at their school means. For details, see @FoxWeisberg:2019 [Sec. 7.2.2].

Still following Fox and Weisberg, we proceed to use the \code{lmer()} function in the \pkg{lme4} package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data:

```{r HSB-lmer, cache=TRUE}
library("lme4", quietly=TRUE)
hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
                   + (cses | school), data=HSB)
summary(hsb.lmer, correlation=FALSE)
```

We can then cross-validate at the cluster (i.e., school) level,

```{r HSB-lmer-CV-cluster, cache=TRUE}
cv(hsb.lmer, k=10, clusterVariables="school", seed=5240)
```

or at the case (i.e., student) level,

```{r HSB-lmer-CV-case, cache=TRUE}
cv(hsb.lmer, seed=1575)
```

For cluster-level CV, the \code{clusterVariables} argument tells \code{cv()} how the clusters are defined. Were there more than one clustering variable, say classes within schools, these would be provided as a character vector of variable names: \code{clusterVariables = c("school", "class")}. For cluster-level CV, the default is \code{k = "loo"}, that is, leave one cluster out at a time; we instead specify \code{k = 10} folds of clusters, each fold therefore comprising $160/10 = 16$ schools.

@@

We start by generating the data, using parameters consistent with the description above and meant to highlight the issues that arise in cross-validating mixed-effects models:[^5]

[^5]: We invite the interested reader to experiment with varying the parameters of our example.

@@

We proceed to fit several mixed-effects models to the data, using the `compareCoefs()` function from the \pkg{car} package [@FoxWeisberg:2019] to display the fixed-effect estimates for these models. We discuss each of these models below.

```{r fitted-mixed-models}
mod.0 <- lmer(y ~ 1 + (1 | group), Data)
mod.1 <- lmer(y ~ x + (1 | group), Data)
mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
library("car", quietly=TRUE)
compareCoefs(mod.0, mod.1, mod.2, mod.3)
```

@@

To illustrate cross-validation of model specification, we return to the \code{Auto} data set:[^6]

[^6]: This example benefits from an email conversation with Bill Venables, who of course isn't responsible for the use to which we've put his insightful remarks.

```{r Auto-redux}
names(Auto)
xtabs(~ year, data=Auto)
xtabs(~ origin, data=Auto)
xtabs(~ cylinders, data=Auto)
```

The \code{Auto} data appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}. Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data. We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore}, produced by the \code{scatterplotMatrix()} function in the \pkg{car} package):

```{r Auto-explore, fig.show="hold"}
#| out.width = "45%",
#| fig.height = 5,
#| fig.cap = "(a) Scatterplot matrix for the numeric variables in the \\code{Auto} data. In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression. (b) Component+residual plots for the working model fit to the \\code{Auto} data. The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.",
#| fig.subcap=rep("", 2)
Auto$cylinders <- factor(Auto$cylinders,
                         labels=c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels=c("America", "Europe", "Japan"))
rownames(Auto) <- make.names(Auto$name, unique=TRUE)
Auto$name <- NULL

scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
                  smooth=list(spread=FALSE), data=Auto, pch=".")

m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch=".")
```

```{r Auto-explore-show, eval=FALSE}
Auto$cylinders <- factor(Auto$cylinders,
                         labels=c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels=c("America", "Europe", "Japan"))
rownames(Auto) <- make.names(Auto$name, unique=TRUE)
Auto$name <- NULL

scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
                  smooth=list(spread=FALSE), data=Auto, pch=".")
```

A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors. There are, however, only 5 distinct values of \code{cylinders} (ranging from 3 to 8), but cars with 3 or 5 cylinders are rare, and none of the cars has 7 cylinders. There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^7] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.

[^7]: Making the decision to treat \code{year} as a factor on this basis could be construed as cheating in the current context, which illustrates the difficulty of automating the whole model-selection process. It's rarely desirable, in our opinion, to forgo exploration of the data to ensure the purity of model validation. We believe, however, that it's still useful to automate as much of the process as we can, as illustrated here, to obtain a more realistic, if still biased, estimate of the predictive power of a model.

We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:

```{r Auto-working-model-show, eval=FALSE}
m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch=".")
```

The component+residual plots in Figure \ref{fig:Auto-working-model} clearly reveal the inadequacy of the model.

@@

Applying `selectModelList()` to the `Auto` polynomial-regression models, and using 10-fold CV, we obtain:

```{r recursive-CV-polynomials}
recursiveCV.auto <- cv(selectModelList, data=Auto, 
                       working.model=mlist,
                       save.model=TRUE, seed=2120)
recursiveCV.auto
recursiveCV.auto$selected.model
cv(mlist[[7]], seed=2120) # same seed for same folds
```

As expected, recursive CV produces a larger estimate of MSE for the selected 7th degree polynomial model than CV applied directly to this model.
